### 2015-04-23 Rmarkdown/comethylation.Rmd
Estimation of odds ratios in 2x2 and 2x2xK tables

Consider two binary random variables, $X$ and $Y$. Let $\pi_{ij}, i, j = 1, 2$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$. We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$. There are many reasons why we might wish to estimate the odds ratio (__CITE__), but our interest here is __WHAT__.

@@

The simplest estimator of the odds ratio is the unconditional maximum likelihood estimator, $\widehat{\psi_{U}} = \frac{n_{11} n_{22}}{n_{12} n_{21}}$.

- Go through the various estimators, their pros and cons.
   - Conditional maximum likelihood estimator
- Explain the advantage of the "0.5-adjusted" unconditional estimator.
- Move to $2 \times 2 \times K$ tables
  - Go through same comparison of estimators
- Move to sparseness
  - Asymptopia I and II
  - Go through same comparison of estimators
  - The homogeneity assumption

@@

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ 0.5 * Gaussian(log2(or_mean), 1) + 0.5 * Gaussian(-log2(or_mean))
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
w <- sample(c(-1, 1), K, replace = TRUE)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, w * log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = NM(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
```

>>>>

We will begin with an overview of the simplest case of estimating dependence in $2 \times 2$ contingency tables, i.e., $K = 1$. We will then progressively introduce the complications of sparsity, $K > 1$, and heterogeneity across the $K$ tables. We conclude with a simulation study highlighting the performance of several difference estimators and determining which is most appropriate for a study of within-fragment co-methylation.

\subsection{$2 \times 2$ contingency tables}

Consider two binary random variables, $X$ and $Y$. Let $\pi_{ij}, i, j = 1, 2$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$. We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$. There are many reasons why we might wish to estimate the odds ratio (__CITE__), but our interest here is __WHAT__. Provided that all $\pi_{ij} > 0$, the odds ratio, $\psi$, can be interpreted as follows:

- $\psi = 1$: $X$ is independent of $Y$.
- $\psi > 1$: Subjects are more likely to have the same level for both $X$ and $Y$, e.g., subjects with $(X, Y) = (1, 1)$ or $(X, Y) = (2, 2)$ are more likely than subjects with $(X, Y) = (1, 2)$ or $(X, Y) = (2, 1)$.
- $\psi < 1$: Subjects are more likely to have the the opposite level for $X$ than they do for $Y$, e.g., subjects with $(X, Y) = (1, 2)$ or $(X, Y) = (2, 1)$ are more likely than subjects with $(X, Y) = (1, 1)$ or $(X, Y) = (2, 2)$.

The possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\phi = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\phi$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

@@

The simplest estimator of $\phi$ is the unconditional maximum likelihood estimator, $\widehat{\phi_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$. The asymptotic standard error of $\widehat{\phi_{U}}$ is given by $\hat{\sigma}(\widehat{\phi_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. The asymptotic distribution of $\widehat{\phi_{U}}$ as $n \to \infty$ is $Normal(\phi, \hat{\sigma}(\phi)^{2})$.

A problem with the estimator $\widehat{\phi_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities. We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\phi_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\phi_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. \citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\phi_{U}}$ and $\widehat{\phi_{0.5}}$ have the same asymptotic distribution around $\phi$ as $n \to \infty$ but that $\widehat{\phi_{0.5}}$ has reduced first-order bias. The modified estimator, $\widehat{\phi_{0.5}}$, is therefore generally recommended.

Another alternative to $\widehat{\phi_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\phi_{C}}$. This is computed by first conditioning on ${n_{1 \cdot}, n_{\cdot 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 \cdot}, n_{\cdot, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{psi_{C}}$ is solved using iterative methods.

The conditional estimator, $\widehat{\phi_{C}}$, works better than the unconditional estimator, $\widehat{\phi_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}. Estimation of the odds ratio is made more difficult when the contingency table is _sparse_. We say that a contingency table is sparse when some of the $n_{ij}$ are small. Sparsity can occur even when the sample size, $n$, is large. Sparsity becomes more of a problem as we move from a single $2 \times 2$ table to $K$ $2 \times 2$ tables or, equivalently, a $2 \times 2 \times K$ table.

\subsection{$2 \times 2 \times K$ contingency tables}

Suppose we now measure the same two binary variables $X$ and $Y$ in $K$ different strata. We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum. For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{\cdot 1k} = 1 - \pi_{\cdot 2k}, \pi_{1 \cdot k} = 1 - \pi_{2 \cdot k}$ to vary.

We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\phi = \log{\psi}$ \citep[see][]{Breslow:1981ta}.

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{\cdot \cdot k})}{n_{12k} n_{21k} / n_{\cdot \cdot k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\phi_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\phi_{MH}})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.

[^mh_estimator]: Other estimators had been proposed but were supplanted by the work of \citet{Robins:1986vh}.

The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{\cdot \cdot k}$, grows large. The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{\cdot \cdot k}$ remains small. The variance estimator proposed by \citet{Robins:1986vh} is consistent under both $A1$ and $A2$. For our study of within-fragment co-methylation, $A2$ is the more relevant asymptotic environment.

\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, converge to the true odds ratio. In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi_{MH}}$, and the conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, do convert to $\psi$, with $\widehat{\psi_{MH}}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.

\subsection{Homogeneity of odds ratios assumption}

So far we have assumed a common odds ratio for the $K$ $2 \times 2$ tables.

__UP TO HERE__

- Move to $2 \times 2 \times K$ tables
  - Go through same comparison of estimators
- Move to sparseness
  - Asymptopia I and II
  - Go through same comparison of estimators
  - The homogeneity assumption

@@

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ 0.5 * Gaussian(log2(or_mean), 1) + 0.5 * Gaussian(-log2(or_mean))
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
w <- sample(c(-1, 1), K, replace = TRUE)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, w * log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = NM(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# TODO: Increase d for each table
```

### 2015-04-23 latex/phd_thesis.tex
Estimation of odds ratios in 2x2 and 2x2xK tables

\chapter{List of Abbreviations}

@@

bp & base pair \\
kb & kilobase = $1000$ bp \\
Mb & megabase = $1,000,000$ bp \\
Gb & gigabase = $1,000,000,000$ bp \\
WGBS & whole genome bisulfite sequencing \\
RRBS & reduced representation bisulfite sequencing \\
FDR & false discovery rate \\
DMR & differentially methylated region \\
DMC & differentially methylated cytosine \\
HMM & hidden Markov model \\
CNV & copy number variant \\
SE & single-end sequencing \\
PE & paired-end sequencing \\
indel & insertion or deletion \\

>>>>

% TODO: Check list of abbreviations
\chapter{List of Abbreviations}

@@

bp & base pair \\
kb & kilobase = $1000$ bp \\
Mb & megabase = $1,000,000$ bp \\
Gb & gigabase = $1,000,000,000$ bp \\
WGBS & whole genome bisulfite sequencing \\
RRBS & reduced representation bisulfite sequencing \\
FDR & false discovery rate \\
DMR & differentially methylated region \\
DMC & differentially methylated cytosine \\
HMM & hidden Markov model \\
SE & single-end sequencing \\
PE & paired-end sequencing \\
indel & insertion or deletion \\
CGI & CpG island \\
